---
title: "Assignment 1"
author: "Samuel Abächerli"
date: "10/31/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem 1: Indexing

Indexing is an important part of working with data. Many people use specialized data bank
programs like Acess or SQL for that. If you are good with R, you don’t need a specialized
program. Import the data file ACC_data.txt into R. Generate a subset of this data as follows:

a) properties located in North Carolina or South Carolina
Hint: an easy solution is the | operator. Feel like getting an extra point? Try the %in%
operator (4 P)

b) The 90% quantile of the property size of those properties that are located in Tallahassee or Athens.
Hint: use the | operator, quantile(), and dat1$size[]. (4 P)

```{r}
# set working directory
# 

# import ACC_data.txt
ACC <- read.table("ACC_data.txt", header = T)

# 1a
# PROPERTIES LOCATED IN NORTH OR SOUTH CAROLINA WITH | OPERATOR
ACC1 <- ACC[ which(ACC$State.Province.Country == "North Carolina, USA" | ACC$State.Province.Country == "South Carolina, USA"),]

# PROPERTIES LOCATED IN NORTH OR SOUTH CAROLINA WITH %IN%
# creating subsets of ACC where the state is either South Caroline and North Carolina
ACC.SouthCarolina <- ACC[ACC$State.Province.Country %in% "South Carolina, USA",]
ACC.NorthCarolina <- ACC[ACC$State.Province.Country %in% "North Carolina, USA",]
# binding the two data frames by row
ACC2 <- rbind(ACC.NorthCarolina, ACC.SouthCarolina)

# 1b
# 90% QUANTILE WITH | OPERATOR
# subsetting ACC where the city is either Athens or Tallahassee
ACC3 <- ACC[ which(ACC$City == "Athens" | ACC$City == "Tallahassee"),]
# calculate the 90 percent quantile with regard to the size
Q90 <- quantile(ACC3$Size, probs = 0.9)
# remove all entries in the data frame that have a size above the 90th percentile (and ordered by size)
ACC3.Q90 <- ACC3[order(ACC3$Size),]
for (row in 1:nrow(ACC3.Q90)) {
  if (ACC3.Q90[row, "Size"] > Q90) {
    ACC3.Q90 <- ACC3.Q90[-row,]
  }
}
```

## Problem 2: tapply()

Import the data file ACC_data.txt into R. Compute the average property size for each U.S.
state. Sort the mean size in decreasing order and plot them using barplot().
Hint: For this problem you need the commands tapply(), sort(), and barplot(). Don’t like the
way bar labels are displayed on the x-axis? Fix the problem and gain an extra point! (7 P)

```{r}
# AVERAGE PROPERTY SIZE OF EACH STATE
# calculate the mean for each state
ACC4 <- tapply(ACC$Size, ACC$State.Province.Country, mean)
# order the data alphabetically
ACC4.ordered <- ACC4[order(names(ACC4))]
# set the names to the abbreviations of the states
names(ACC4.ordered) <- c("AL", "AZ", "CA", "CO", "FL", "GA", "IL", "IN", "IA", "KY", "LA", "MI", "MN", "MS", "NE", "NV", "NJ", "NY", "NC", "OH", "OK", "PA", "SC", "TN", "TX", "UNDEF", "VA", "WV")
# sort the data in decreasing order
ACC4.sorted <- sort(ACC4.ordered, decreasing = T)
```

## Plots

Average property size of each state in decreasing order with abbreviations as labels.

```{r ACC4.sorted}
# plot the data in a bar plot with vertical labels
barplot(ACC4.sorted, las = 2)
```

